Advances in Artificial Intelligence for Data Visualization: Developing Computer Vision Models to Automate Reading of Data Plots, with Application to Predictive Model Diagnostics

### 2022-03-03 Thesis/01-chap1.Rmd
comments up to 1.6

As discussed in @buja_statistical_2009, in the practice of model diagnostics, the range of possible visual discoveries is not pre-specified. In other words, people do not explicitly specify which one or more visual features they are looking for before the read of the diagnostic plot. This is concerning since conventional hypothesis testing always requires the pre-specification of the parameter space $\Theta$ of the parameter of interest $\theta \in \Theta$ to form a valid inferential procedure. To address this issue, a collection of test statistics $T^{(i)}(\boldsymbol{\mathrm{y}})~(i \in I)$ is defined, where $\boldsymbol{\mathrm{y}}$ is the data and $\boldsymbol{I}$ is a set of all possible visual features. @buja_statistical_2009 described each of the test statistics $T^{(i)}(\boldsymbol{\mathrm{y}})$ as a measurement of the degree of presence of a visual feature. Alternatively, @majumder_validation_2013 avoids the use of visual features and defined the visual statistics $T(.)$ as a mapping from a dataset to a data plot. Both definitions of visual test statistics are valid, but in the rest of the paper the first definition will be used as it covers some details needed by the following discussion.

@@

@buja_statistical_2009 proposed the lineup protocol as a visual test to calibrate the Type I error issue without the specification of $c^{(i)}~(i \in I)$. It is inspired by the "police lineup" or "identity parade" which is the act of asking the eyewitness to identify criminal suspect from a group of irrelevant people. The protocol consists of $m$ randomly placed data plots, where $1$ plot is the actual data plot, and $m-1$ null plots are produced by plotting data simulate from the null distribution which is consistent with the null hypothesis. Then, an observer who have not seen the actual data plot will be asked to point out the most different plot from the lineup.

Under the null hypothesis, it is expected that the actual data plot would have no distinguishable difference with the null plots, and the probability of the observer correctly picks the actual data plot is $1/m$ due to randomness. If we reject the null hypothesis as the observer correctly picks the actual data plot, then the Type I error of this test is $1/m$.

This provides us with an mechanism to control the Type I error, because $m$ - the number of plots in a lineup can be chosen. A larger value of $m$ will result in a smaller Type I error, but the limit to the value of $m$ depends on the number of plots a human willing to view [@buja_statistical_2009]. Typically, $m$ will be set to $20$ which is equivalent to set $\alpha = 0.05$, a general choice of significance level for conventional testing among statisticians.

@@

@@

Assumes that individuals have the same ability and a lineup has been evaluated by multiple individuals, under the alternative hypothesis, the estimated power for a lineup can be expressed as $\hat{p} = x/K$, the estimated probability of identifying the actual data plot from the lineup. If the individual skill needs to be taken into account, and $L$ lienups have been evaluated by $K$ individuals, @majumder_validation_2013 suggests that mixed effects logistic regression model can be fit as:

$$g(p_{li}) = W_{li}\delta + Z_{li}\tau_{li},$$
where $g(.)$ is the logit link function $g(p) = log(p)  - log(1-p)$; $0 \leq p \leq 1$. $W_{li}$, $1 \leq i \leq K$, $1 \leq l \leq L$, is the covariate matrix including lineup-specific elements and demographic information of individuals, and $\delta$ is a vector of parameters. $Z$ is the random effects matrix, and $\tau$ is a vector of variables follow $N(\boldsymbol{0},\sigma_{\tau}\boldsymbol{I}_{KL\times KL})$.

>>>>

As discussed in @buja_statistical_2009, in the practice of model diagnostics, the range of possible visual discoveries is not pre-specified. In other words, people do not explicitly specify which visual feature(s) they are looking for before the reading the diagnostic plot. In contrary, conventional hypothesis testing always requires the pre-specification of the parameter space $\Theta$ of the parameter of interest $\theta \in \Theta$ to form a valid inferential procedure. To address this issue, a collection of test statistics $T^{(i)}(\boldsymbol{\mathrm{y}})~(i \in I)$ is defined, where $\boldsymbol{\mathrm{y}}$ is the data and $I$ is a set of all possible visual features. @buja_statistical_2009 described each of the test statistics $T^{(i)}(\boldsymbol{\mathrm{y}})$ as a measurement of the degree of presence of a visual feature. Alternatively, @majumder_validation_2013 avoids the use of visual features and defined the visual statistics $T(.)$ as a mapping from a dataset to a data plot. Both definitions of visual test statistics are valid, but in the rest of the paper the first definition will be used as it covers some details needed by the following discussion.

@@

@buja_statistical_2009 proposed the lineup protocol as a visual test to calibrate the Type I error issue without the specification of $c^{(i)}~(i \in I)$. It is inspired by the "police lineup" or "identity parade" which is the act of asking the eyewitness to identify criminal suspect from a group of irrelevant people. The protocol consists of $m$ randomly placed plots, where one plot is the actual data plot, and the remaining $m - 1$ plots have the identical graphical production as the data plot except the data has been replaced with data consistent with the null hypothesis. Then, an observer who have not seen the actual data plot will be asked to point out the most different plot from the lineup.

Under the null hypothesis, it is expected that the actual data plot would have no distinguishable difference with the null plots, and the probability of the observer correctly picks the actual data plot is $1/m$. If we reject the null hypothesis as the observer correctly picks the actual data plot, then the Type I error of this test is $1/m$.

This provides us with an mechanism to control the Type I error, because $m$ - the number of plots in a lineup can be chosen. A larger value of $m$ will result in a smaller Type I error, but the limit to the value of $m$ depends on the number of plots a human is willing to view [@buja_statistical_2009]. Typically, $m$ will be set to $20$ which is equivalent to set $\alpha = 0.05$, a general choice of significance level for conventional testing among statisticians.

@@

[ET: Multiple observers, to me, is not about making smaller p-value possible with finite $m$ but if you only have a single observer, you are confounding the test with the visual ability of that single observer. In some ways, you can consider multiple observers as an ensemble approach and the decision is made on pooling the results from multiple "methods", each with their own abilities.]

@@

Suppose individuals have the same ability and a lineup has been evaluated by multiple individuals, under the alternative hypothesis, the estimated power for a lineup can be expressed as $\hat{p} = x/K$, the estimated probability of identifying the actual data plot from the lineup. If the individual skill needs to be taken into account, and $L$ lineups have been evaluated by $K$ individuals, @majumder_validation_2013 suggests that mixed effects logistic regression model can be fit as:

$$g(p_{li}) = W_{li}\delta + Z_{li}\tau_{li},$$
where $g(.)$ is the logit link function $g(p) = log(p)  - log(1-p)$; $0 \leq p \leq 1$. $W_{li}$, $1 \leq i \leq K$, $1 \leq l \leq L$, is the covariate matrix including lineup-specific elements and demographic information of individuals, and $\delta$ is a vector of fixed effects. $Z_{li}$ is the random effects matrix, and $\tau ~ N(\boldsymbol{0}_{KL},\sigma_{\tau}\boldsymbol{I}_{KL\times KL})$.

[ET: need to double check this. Is $W_{li}$ a matrix or an element of the matrix? I wouldn't say "$Z_{li}$ is the random effects matrix". It's gotta be like random intercept model for lineup id and another random intercept for person id? ET to check.]
